---
title: 'Data Processing in R: <br/> Or, how to rid your workflow of Excel'
author: "Jessica Luo"
date: "March 10, 2015"
output: ioslides_presentation
---

## A typical workflow

- Acquire raw data
- Open with Excel
  - Examine the data, acquire some summary stats, generate initial plots
  - Manipulate the spreadsheet, save "version_1.xls"
- Import into stats software, performs stats


- But you can do all of this in R!

## Getting started

```{r}
# load libraries
library(ggplot2)
library(plyr)
library(stringr)
suppressWarnings(library(reshape2))
suppressMessages(library(oce))
# used these 'suppress' commands for a smoother presentation

# read in dataset
d <- read.csv("data/CAAL_Incubation/27_405_90186_201113.csv", 
              header=T, stringsAsFactors=F)
```
- Note that you can use read.csv or read.table

## Examining initial data frame

```{r}
head(d)
```

## Summary stats

```{r}
# can use range(), mean(), and summary() to examine your dataset
range(d$Date)
summary(d)
```

## Convert to dateTime format - POSIX

```{r}
# first use string concatenate to bind the strings together
# str_c() is from the stringr package
d$dateTime <- str_c(d$Date, d$Time)
head(d$dateTime)
# note that str_c() is not the same as c()
```


## Convert to dateTime format - POSIX

```{r}
# POSIX format will allow you to manipulate the dateTime
?as.POSIXct
d$dateTime <- as.POSIXct(d$dateTime, format="%Y/%m/%d %H:%M:%S", tz="UTC")
head(d$dateTime)
# can also calculate a "day of year"
d$doy <- format(d$dateTime, "%j")
head(d$doy)
```


## Renaming & removing columns

```{r}
# easy renaming from plyr package
d <- rename(d, replace=c("Latitude"="lat", "Longitude"="lon"))
names(d)
# select columns to keep using subsetting and the c() command
d <- d[,c("dateTime", "lat", "lon", "doy")]
head(d)
```

## Calculating a distance measure
```{r}
?geodDist # from the oce package, dist in km
# assume that the first lat, lon is the colony location
d$dist_col <- geodDist(d$lat, d$lon, d$lat[1], d$lon[1])
# note that we created a new column just by calling it
```
Alternatively -- can write a function:
```{r}
# Compute the straight line distance from the starting point of a lat,lon trajectory
dist.from.start <- function(lat, lon) {
  library("oce")
  geodDist(lat1=lat, lon1=lon, lat2=na.omit(lat)[1], 
           lon2=na.omit(lon)[1], alongPath=FALSE)
}
d$dist_col <- dist.from.start(d$lat, d$lon)
head(d$dist_col)
```

## Examining the distance traveled

```{r, fig.show='asis'}
# initial plot (note: fig not cut off in R console)
ggplot() + geom_point(aes(x=dateTime, y=dist_col), data=d)
```

## Keep only data where dist > 5 km

```{r}
d$speed <- NA
d$trip_dur <- NA
# use the which command to find the indices (row #s)
# where the distance is over 5 km
distOver5km <- which(d$dist_col >= 5)
length(distOver5km) # how many indices
head(distOver5km) # where does it start
tail(distOver5km) # where does it end
```

## Keep only data where dist > 5 km

```{r}
first <- distOver5km[1]
last <- distOver5km[length(distOver5km)]    

# subsetting rows
d <- d[first:last,]
head(d)
```

## Keep only data where dist > 5 km

```{r}
first <- distOver5km[1]
last <- distOver5km[length(distOver5km)]    

# subsetting rows
d <- d[first:last,]
head(d)
```

## Calculate speed

```{r}
dist <- geodDist(d$lat, d$lon, alongPath=TRUE) # distance along a path   
# diff() calculates a difference in value between two consecutive points
# or you can set different lag values as well. default lag=1
diffdist <- diff(dist) 
head(diffdist)
length(dist)
length(diffdist)
```

## Calculate speed

```{r}
diff_time <- diff(d$dateTime) # distance along a path   
head(diff_time) # still a POSIX format
diff_time <- as.numeric(diff_time, units="hours") # change to numeric
head(diff_time)
speed <- c(0, diffdist/diff_time) # need to add a 0 to the beginning
# now is the same length as our data frame
d$speed <- speed
```

## Calculate trip duration
```{r}
?difftime # calculates a time difference b/t two times
# for example:
difftime(d$dateTime[2], d$dateTime[1])
# also works on arrays
difftime(d$dateTime[2:10], d$dateTime[1], units="hours")

# so can apply this across the whole data frame
d$trip_dur <- difftime(d$dateTime, d$dateTime[1], units = "hours")   
# Note that none of this works if there are no trips over 5 km
```

## Use a conditional statement
```{r}
    if (length(distOver5km) != 0){
      first <- distOver5km[1]
      last <- distOver5km[length(distOver5km)]     
      d <- d[first:last,]
      # speed in km / hr
      speed <- diff(geodDist(d$lat, d$lon, alongPath=TRUE))/
        as.numeric(diff(d$dateTime), units="hours")
      d$speed <- c(0, speed) 
      d$trip_dur <- difftime(d$dateTime, d$dateTime[1], units = "hours")    
    }

  if(length(distOver5km) == 0){
    # return an empty data fram with the same names and dimensions
    names <- names(d)
    # this starts a new blank data frame from an empty matrix
    d <- data.frame(matrix(NA, ncol=length(names(d)), nrow=1))
    names(d) <- names
    } 
```

## And now, repeat x 20, or x 200
To repeat this process there are a couple different ways:

> *for()* loops:  
- Read in names for each of the files  
- Loop through the names, and apply the same procedure to each  
- Bind all the rows together at each step <br>   
> *plyr* package has *adply()*  
- *plyr* uses the split-apply-combine strategy  
- For this problem, we are splitting the "list" (array) of filenames  
- Apply a function (*read.csv*, and all other processes)  
- Then combine it all together

## Demonstration with *adply()*
```{r}
# load data
raw_data <- "data/CAAL_incubation"
files <- list.files(raw_data, full=T)

d <- adply(files, 1, function(file){
  d <- read.csv(file)
  # pull the band number out of the file name
  d$band <- str_sub(file, -16, -12)

  d <- rename(d, replace=c("Latitude" = "lat", "Longitude"="lon"))
  
  d$lonn <- ifelse(d$lon < 0, d$lon + 360, d$lon)
  
  # create a proper datetime
  d$dateTime <- as.POSIXct(strptime(str_c(d$Date, d$Time), format="%Y/%m/%d %H:%M:%S", tz="UTC"))
  
  # "day of year" and "time of day" becomes unnecesary with a proper dateTime that R can recognize and manipulate, but we can still calculate these terms
  d$doy <- format(d$dateTime, "%j")
  # time of day a little harder to calculate so leaving it for now
  
  d$dist_col <- dist.from.start(d$lat, d$lon)
  d <- d[, c("dateTime", "lat", "lon", "lonn", "band", "dist_col", "doy")]
  d$speed <- NA
  d$trip_dur <- NA
  # cut off points - trips are defined as anything > 5 km from nest site
  # demonstrate usage of "which"
  temp <- which(d$dist_col >= 5)
  if (length(temp) != 0){
    first <- temp[1]
    last <- temp[length(temp)]    
    # subsetting
    d <- d[first:last,]
    
    # speed in km / hr
    speed <- diff(geodDist(d$lat, d$lon, alongPath=TRUE))/as.numeric(diff(d$dateTime), units="hours")
    d$speed <- c(0, speed) 
    
    d$trip_dur <- difftime(d$dateTime, d$dateTime[1], units = "hours")    
    }
  if(length(temp) == 0){
    names <- names(d)
    d <- data.frame(matrix(NA, ncol=length(names(d)), nrow=1))
    names(d) <- names} 
  
  return(d)
  
}, .progress="text")

# remove adply column
d <- d[,-1]
d <- d[complete.cases(d),]
row.names(d) <- 1:nrow(d)

d$dateTime <- as.POSIXct(d$dateTime, origin = "1970-01-01 00:00.00", tz="UTC")
write.csv(d, "CAAL_incubation_set1.csv", row.names=F)
```

## Demonstration with *for()* loops
```{r}
# load data
raw_data <- "data/CAAL_incubation"
files <- list.files(raw_data, full=T)

d <- NULL

for (i in 1:length(files)){
  x <- read.csv(files[i])
  # pull the band number out of the file name
  x$band <- str_sub(files[i], -16, -12)
  
  names(x)[3] <- "lat"
  names(x)[4] <- "lon"
  
  x$lonn <- ifelse(x$lon < 0, x$lon + 360, x$lon)
  
  # create a proper datetime
  x$dateTime <- as.POSIXct(strptime(str_c(x$Date, x$Time), format="%Y/%m/%d %H:%M:%S", tz="UTC"))
  
  # "day of year" and "time of day" becomes unnecesary with a proper dateTime that R can recognize and manipulate, but we can still calculate these terms
  x$doy <- format(x$dateTime, "%j")
  # time of day a little harder to calculate so leaving it for now
  
  x$dist_col <- dist.from.start(x$lat, x$lon)
  x <- x[, c("dateTime", "lat", "lon", "lonn", "band", "dist_col", "doy")]
  x$speed <- NA
  x$trip_dur <- NA
  # cut off points - trips are defined as anything > 5 km from nest site
  # demonstrate usage of "which"
  temp <- which(x$dist_col >= 5)
  if (length(temp) != 0){
    first <- temp[1]
    last <- temp[length(temp)]    
    # subsetting
    x <- x[first:last,]
    
    # speed in km / hr
    speed <- diff(geodDist(x$lat, x$lon, alongPath=TRUE))/as.numeric(diff(x$dateTime), units="hours")
    x$speed <- c(0, speed) 
    
    x$trip_dur <- difftime(x$dateTime, x$dateTime[1], units = "hours")    
  }
  if(length(temp) == 0){
    names <- names(x)
    x <- data.frame(matrix(NA, ncol=length(names(x)), nrow=1))
    names(x) <- names} 

  # bind together
  d <- rbind(d, x)
}

d <- d[complete.cases(d),]
row.names(d) <- 1:nrow(d)

d$dateTime <- as.POSIXct(d$dateTime, origin = "1970-01-01 00:00.00", tz="UTC")

write.csv(d, "CAAL_incubation_set1.csv", row.names=F)
```


## A few nifty *dplyr* functions
```{r, tidy=T}
suppressMessages(library(dplyr))

d <- tbl_df(d) # dplyr format, data frame table
# easy way to quickly look at d without typing (head(d))
d 
```

## A few nifty *dplyr* functions
```{r}
# dplyr summarize for summary stats
d %>% group_by(band) %>% summarize(counts = n(), 
      mean.dcol = mean(dist_col), mean.spd = mean(speed))
```

## A few ggplot2 plots
```{r, echo=F}
d <- data.frame(d)
data(coastlineWorld) # oce package
coastline.world=data.frame(lon=coastlineWorld[["longitude"]],lat=coastlineWorld[["latitude"]]) 
coastline.world$lonn <- ifelse(coastline.world$lon < 0, coastline.world$lon+360 , coastline.world$lon)
```
```{r}
suppressWarnings(ggplot(mapping=aes(x=lonn, y=lat)) + geom_point(aes(colour=band), data=d) + geom_path(data=coastline.world) + theme_bw() + scale_x_continuous(limits=c(100, 220)) + scale_y_continuous(limits=c(-80, 0)))
```

## A few ggplot2 plots
```{r}
suppressWarnings(ggplot(mapping=aes(x=lonn, y=lat)) + geom_point(aes(alpha=rev(speed), size=rev(speed)), data=d) + geom_path(data=coastline.world) + theme_bw() + scale_x_continuous(limits=c(100, 220)) + scale_y_continuous(limits=c(-80, 0)))
```